---
output:
  pdf_document: default
  html_document: default
---


#  End Member Mixing Analysis
#### Performed with EMMAgeo in R
#### *Author: Andrew Macumber*


## Table of Contents
* [Prep Workspace](#workspace)
* [Load Data](#data)
* [Data Wrangling](#wrangle)
* [Data Objects](#objects)
* [Graph: Grain Size Frequency Distribution](#psagsd)
* [EMMA: Model Parameters](#emmaparameters)
* [EMMA: Run All Models](#emmamodelall)
* [EMMA: Run Specific Model](#emmamodelone)
* [EMMA: Run Robust Model](#emmamodelrobust)
* [EMMA: Results](#emmaresults)
* [Save Data](#save)
* [Graph: Stacked Stratigraphic Plot](#stacked)
* [Graph: Robust End Member Distributions](#emgsd)

## <a name="workspace"></a>Prep Workspace
Remove previous lists from R
```{r}
rm(list=ls())
```

Load libraries
```{r}
library("EMMAgeo")
library("functional")
```

## <a name="data"></a>Load Data
Bring in the dataset
```{r}
psa_v000 <- read.table('HGC2_Shallow_PSA.csv', sep = ",")
```
Check dataframe (optional)
```{r eval = FALSE}
dim(psa_v000)
head(psa_v000)
```

## <a name="wrangle"></a>Data Wrangling
Remove one row and two column headers
```{r}
psa_v001 <- psa_v000[-1, -c(1,2)]
```
Check for grain size bins (columns) with zero sum
```{r}
colSums(psa_v001)
```
Remove NULL values
```{r}
psa_v002 <- na.omit(psa_v001)
```
Convert to double matrix
```{r}
psa_v003 <- as.matrix(psa_v002)
```
Ensure rows sum to 100
```{r}
psa_v004 <- psa_v003 / apply(psa_v003,1,sum) * 100
```
Check dataframe (optional)
```{r eval = FALSE}
dim(psa_v004)
head(psa_v004)
```

## <a name="objects"></a>Data Objects
Create a grain size vector
```{r}
mm_bins <- as.numeric(psa_v000[1,-c(1,2)])
```
Create a depth vector
```{r}
depth <- as.vector(psa_v000[-1,1])
```
Assign an emma object
```{r}
emma <- psa_v004
```

## <a name="psagsd"></a>Graph: Grain Size Frequency Distribution
*Generate a plot of median, minimum and maximum grain size (log) versus frequency (%)*

Create a dataframe with grain size (column) mins, max and medians
```{r}
library("matrixStats")
psa.range <- cbind(mm_bins,
                   colMins(emma, na.rm=TRUE),
                   colMaxs(emma, na.rm=TRUE),
                   colMedians(emma, na.rm=TRUE))
```
Create a graph showing the median, minimum and maximum grain size frequencies
```{r}
plot(psa.range[,1],psa.range[,4], type ="l", log = "x", # plot grain-size (log) versus median
     xlim = c(0.1,2000), ylim = c(0,5), # set the axes limits
     col ="black", xaxt = "n", # colour of the line and suppress x-axis
     xlab = "Grain-size (microns)",
     ylab = "Frequency (%)",
     main = "HGC2 (Shallow Site) Grain-size Distributions") # labels
lines(psa.range[,1],psa.range[,2], col = "black", lty = 2) # add a line for minimum (grey-dash)
lines(psa.range[,1],psa.range[,3], col = "black", lty = 2) # add a line for maximum (grey-dash)
axis (side=1, at = c(0.1,1,10,100,1000)) # add major tick marks
axis (side=1, at = c(seq(0.2,0.9,by = 0.1), seq(2,9, by =1), # add minor tick marks
                     seq(20,90, by = 10), seq(200,900,by=100)),
      labels = NA, tcl=-0.25, lwd=0, lwd.ticks=1) # no labels and shorten tick marks
legend(0.1,4.5,c("Median","Minimum","Maximum"), # add a legend at specific spot with labels
       lty=c(1,2,2), col=c("black","black","black")) # ensure same as lines in plot
```

## <a name="emmaparameters"></a>EMMA: Model Parameters
Check your dataset for errors.
```{r}
check.data(X = emma , q = 10, l = seq(0, 1, 0.01), c = 100)
```
Create a sequence of weight transformations (l)
```{r}
l <- get.l(X = emma, n = 10, max = 0.95, min = 0)
```
Combine the range of end members and weight transformations (Q). Create a range of end members only (q)
```{r}
Q <- get.q(X = emma, l = l)
q <- 2:6
```
See a graph showing how weight transformation (l) varies with number of end Members
```{r}
EM.tp <- test.parameters(X = emma, q = q, l = l, plot = "mRt",
                      legend = "bottomright", cex = 0.7,
                      multicore = FALSE,
                      colour = rgb((1:7) / 7, 0.9, 0.2, 1)
                      )
```
See a table showing how weight transformation (l) varies with number of end-members
```{r}
EM.tp$mRt
```

## <a name="emmamodelall"></a>EMMA: Run All Models
```{r}
EM.pot <- model.EM(X = emma, q = Q)
```

## <a name="emmamodelone"></a>EMMA: Run Specific Model
```{r}
EM.opt <- EMMA(X = emma,
               q = 3, l = 0.009,
               #q = 3, l = 0.012,
               c = 100,
               plot = TRUE,
               classunits = mm_bins,
               ID = as.numeric(depth),
               xlab = c(expression(paste("Class [", micron, "]")), "Depth (cm)"))
```

## <a name="emmamodelrobust"></a>EMMA: Run Robust Model
Get the bin size ranges for end members (i.e., limits)
```{r}
# Visualize limits
stem(EM.pot$modes, scale = 2)

# Assign limits
v1 <- c(28,46,54)
v2 <- c(30,48,56)
limits = cbind(v1,v2)
```
Run robust model
```{r}
EM.rob <- robust.EM(em = EM.pot
                    , l = 0.009
                    , limits = limits
                    , c = 100
                    , plot = TRUE
                    , classunits = mm_bins
                    , ID = as.numeric(depth)
                    , xlab = c(expression(paste("Class [", micron, "]")), "Depth (cm)")
                    )
```

## <a name="emmaresults"></a>EMMA: Results
We tested `r nrow(EM.pot$Vqsn)` models to find the initial parameters. The mean total explained variance:

* For all models: `r max(EM.pot$mRt)`
* For the optimal model: `r EM.opt$mRt`
* For the robust model: `r EM.rob$mRt`

The optimal model had the following end member modes: `r EM.opt$modes` each explaining `r EM.opt$Mqs.var` of the variance, respectively.

The robust model had the following end member modes: `r mm_bins[EM.rob$modes[1]]`, `r mm_bins[EM.rob$modes[2]]`, and `r mm_bins[EM.rob$modes[3]]`. Each explaining `r EM.rob$Mqs.var` of the variance, respectively.

## <a name="save"></a>Save Data
Create a dataframe with robust end member mean abundance and standard deviation.
```{r}
EM.rob.mean <- EM.rob$scores$mean
EM.rob.sd <- EM.rob$scores$sd
EM.rob.scores <- cbind(EM.rob.mean, EM.rob.sd)
colnames(EM.rob.scores) <- c("EM_05_Mean", "EM_27_Mean", "EM_58_Mean",
                             "EM_05_SD", "EM_27_SD", "EM_58_SD")
row.names(EM.rob.scores) <- depth
```
Export File
```{r}
write.csv(EM.rob.scores, file ="HGC2_EMMA_Q3_scores.csv")
```

## <a name="stacked"></a>Graph: Stacked Stratigraphic Plot
Create a stacked stratigraphic plot of robust end member abundance. You will need to format a dataset using "HGC2_EMMA_Q3_scores.csv" that confirms to three columns T, Var, Val:
* T - is depth or time
* Var - is the variable, that is the end member
* Val - is the value or abundance of that end member at that depth or time

For example:\
T VAR VAL\
0 EM01 0.25\
0 EM02 0.50\
0 EM03 0.25\
1 EM01 0.10\
2 EM02 0.85\
3 EM03 0.05

Load Libraries & Data
```{r}
library(ggplot2)
d <- read.table('HGC2_EMMA_Q3_StackedInput_Age.csv', sep = ",", header = TRUE)
```
Create the plot:
1. Create the area plot
2. Change breaks along the x-axis
3. Change the axes labels
4. Change legend labels and make grayscale

```{r}
sa1 <- ggplot(d,aes(x=T,y=VAL,group=VAR,fill=VAR)) + geom_area(position="fill")
sa2 <- sa1 + scale_x_continuous(breaks=seq(1420,2020,50)) + scale_y_continuous(breaks=seq(0,1,1/10))
sa3 <- sa2 + labs(title="Harvey Lake HGC2 Core: End Member Time Series",
                  x="Age (AD)",
                  y="End Member Proportion (%)")
sa4 <- sa3 + scale_fill_grey(name = "End-Members", labels = c("Very Coarse Silt", "Coarse Silt", "Fine Silt")) + theme_bw()
```
View Plot
```{r}
sa4
```

## <a name="emgsd"></a>Graph: Robust End Member Distributions
This figure will show the average grain size distribution with standard deviations for each robust end member. Each end member will be plotted on its on graph. The grain size distributions of all samples will be displayed as grey lines on each graph. *Note: Creates Figure 3 of Chapter 2. Markdown creates an improperly formatted graph. Not run here*

Format sample data
```{r}
dietze <- rbind(mm_bins, emma)
dietze <- t(dietze)
```
Format robust end member means
```{r}
robust <- rbind(mm_bins, EM.rob$loadings$mean)
robust <- t(robust)
```
Format robust end member standard deviations
```{r}
robust.sd <- t(EM.rob$loadings$sd)
```
Plot the graph
```{r eval = FALSE}
plot.new()

# Design the layout of the graph
layout(matrix(c(1,1,2,2,3,3), 3, 2, byrow = TRUE),
       widths=c(3,3,3), heights=c(2,2,2))

# First Graph

## Samples
plot(dietze[,1],dietze[,2], type = "l", log ="x", ylim = c(0,10)
     , xlim = c(0.1,2000),col="grey"
     , xaxt="n", xlab = ""
     , ylab = "Frequency (%)")
for (i in 3:ncol(dietze)) lines(dietze[,1],dietze[,i],col="grey") # plot for every column

## The End Member
lines(robust[,1],robust[,2],type="l",col="black",lty = 1, lwd = 2) # mean
lines(robust[,1],robust[,2]-robust.sd[,1],type="l",col="black",lty = 2, lwd = 2) # lower standard deviation
lines(robust[,1],robust[,2]+robust.sd[,1],type="l",col="black",lty = 2, lwd = 2)

## Graph Details
axis (side=1, at = c(0.1,1,10,100,1000), labels = NA) # add major tick marks
axis (side=1, at = c(seq(0.2,0.9,by = 0.1), seq(2,9, by =1), # add minor tick marks
                     seq(20,90, by = 10), seq(200,900,by=100))
      , labels = NA, tcl=-0.25, lwd=0, lwd.ticks=1) # no labels and shorten tick marks
legend(200,9,c("Samples","End Member Mean","End Member St. Deviation"), # add a legend at specific spot with labels
       lty=c(1,1,2), col=c("grey","black","black")) # ensure same as lines in plot
text(0.1, 8, adj = 0, "Fine Silt End Member (5 microns) Explaining 21% of Variance", cex = 1.5)
title(main = "Robust End Members of Harvey Lake Shallow Core (HGC2)", cex.main = 2)


# Second Graph

## Samples
plot(dietze[,1],dietze[,2], type = "l", log ="x", ylim = c(0,10),
     xlim = c(0.1,2000),col="grey", xaxt="n", xlab = "",
     ylab = "Frequency (%)")
for (i in 3:ncol(dietze)) lines(dietze[,1],dietze[,i],col="grey") # plot for every column

## The End Member
lines(robust[,1],robust[,3],type="l",col="black", lty = 1, lwd = 2)
lines(robust[,1],robust[,3]-robust.sd[,2],type="l",col="black", lty = 2, lwd = 2) # upper standard deviation
lines(robust[,1],robust[,3]+robust.sd[,2],type="l",col="black", lty = 2, lwd = 2)

## Graph Details
axis (side=1, at = c(0.1,1,10,100,1000), labels = NA) # add major tick marks
axis (side=1, at = c(seq(0.2,0.9,by = 0.1), seq(2,9, by =1), # add minor tick marks
                     seq(20,90, by = 10), seq(200,900,by=100))
      , labels = NA, tcl=-0.25, lwd=0, lwd.ticks=1) # no labels and shorten tick marks
legend(200,9,c("Samples","End Member Mean","End Member St. Deviation"), # add a legend at specific spot with labels
       lty=c(1,1,2), col=c("grey","black","black")) # ensure same as lines in plot
text(0.1, 8, adj = 0, "Coarse Silt End Member (27 microns) Explaining 42% of Variance", cex = 1.5)


# Third Graph

## Samples
plot(dietze[,1],dietze[,2], type = "l", log ="x", ylim = c(0,10),
     xlim = c(0.1,2000),col="grey", xaxt="n", xlab = "",
     ylab = "Frequency (%)")
for (i in 3:ncol(dietze)) lines(dietze[,1],dietze[,i],col="grey") # plot for every column

## The End Member
lines(robust[,1],robust[,4],type="l",col="black",lty = 1, lwd = 2)
lines(robust[,1],robust[,4]-robust.sd[,3],type="l",col="black",lty = 2, lwd = 2)
lines(robust[,1],robust[,4]+robust.sd[,3],type="l",col="black",lty = 2, lwd = 2)

## Graph Details
axis (side=1, at = c(0.1,1,10,100,1000)) # add major tick marks
axis (side=1, at = c(seq(0.2,0.9,by = 0.1), seq(2,9, by =1), # add minor tick marks
                     seq(20,90, by = 10), seq(200,900,by=100))
      , labels = NA, tcl=-0.25, lwd=0, lwd.ticks=1) # no labels and shorten tick marks
legend(200,9,c("Samples","End Member Mean","End Member St. Deviation"), # add a legend at specific spot with labels
       lty=c(1,1,2), col=c("grey","black","black")) # ensure same as lines in plot
text(0.1, 8, adj = 0, "Very Coarse Silt End Member (58 microns) Explaining 33% of Variance", cex = 1.5)
title(xlab = "Grain size (microns)")
```
